---
title: "Title of Your Report"
author: "Names of your Group Members"
date: "Due Date"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


# Title of your Report

## Name(s) of Author(s) 
## Date


# Model



## Model Specifics

pursue three logistic models: Trump or not Trump, Biden or not Biden, Trump or Biden
$$ log$$

Where $y$ represents the proportion of voters who will vote for Donald Trump. Similarly, $\beta_0$ represents the intercept of the model, and is the probability of voting for Donald Trump at age 0. Additionally, $\beta_1$ represents the slope of the model. So, for everyone one unit increase in age, we expect a $\beta_1$ increase in the probability of voting for Donald Trump.

#Trump or not trump

```{r, include=FALSE}
library(tidyverse)
setwd("/Users/ZHANGMIN/Desktop")

survey_data <- read_csv("survey_data.csv")
census_data <- read_csv("census_data.csv")

# Creating the Model: Trump or not Trump
model_1 <- glm(vote_trump ~ age+as.factor(sex)+as.factor(race)+as.factor(education), 
            data=survey_data, family= "binomial")
summary(model_1)    # age >= 18

# OR
# broom::tidy(model)

```

## Post-Stratification 

In order to estimate the proportion of voters who will vote for Donald Trump I need to perform a post-stratification analysis. Here I create cells based off different ages, race, eduction level. Using the model described in the previous sub-section I will estimate the proportion of voters in each age bin. I will then weight each proportion estimate (within each bin) by the respective population size of that bin and sum those values and divide that by the entire population size. 

```{r, include=TRUE}

# Here I will perform the post-stratification calculation
census_data$logodds_estimate <-
  model_1 %>%
  predict(newdata = census_data)

census_data$estimate <-
  exp(census_data$logodds_estimate)/(1+exp(census_data$logodds_estimate))

census_data %>%
  mutate(trump_predict_prop = estimate*n) %>% 
  summarise(trump_predict = sum(trump_predict_prop)/sum(n))

```


#biden or not biden

```{r, include=FALSE}
library(tidyverse)
setwd("/Users/ZHANGMIN/Desktop")
survey_data <- read_csv("survey_data.csv")
census_data <- read_csv("census_data.csv")

# Creating the Model: Biden or not Biden
model_2 <- glm(vote_biden ~ age+as.factor(sex)+as.factor(race)+as.factor(education), 
            data=survey_data, family= "binomial")
summary(model_2)    # age >= 18
```

```{r, include=FALSE}
census_data$logodds_est <-
  model_2 %>%
  predict(newdata = census_data)

census_data$estimate <-
  exp(census_data$logodds_est)/(1+exp(census_data$logodds_est))

census_data <-
  census_data %>% 
  mutate(biden_predict_prop = estimate*n) %>% 
  summarise(biden_predict = sum(biden_predict_prop)/sum(n))
view(census_data)
```

#Trump or Biden: 


```{r, include=FALSE}
library(tidyverse)
setwd("/Users/ZHANGMIN/Desktop")
survey_data <- read_csv("survey_data_2.csv")
census_data <- read_csv("census_data.csv")

# Creating the Model: Biden or not Biden
model_3 <- glm(biden_or_trump ~ age+as.factor(sex)+as.factor(race)+as.factor(education), 
            data=survey_data, family= "binomial")
summary(model_3)    # age >= 18
```

```{r, include=FALSE}
census_data$logodds_est <-
  model_3 %>%
  predict(newdata = census_data)

census_data$estimate <-
  exp(census_data$logodds_est)/(1+exp(census_data$logodds_est))

election_result <-
  census_data %>% 
  mutate(biden_win = estimate*n) %>% 
  summarise(biden_or_trump = sum(biden_win)/sum(n))
view(election_result)
```



















# Results

Here you will include all results. This includes descriptive statistics, graphs, figures, tables, and model results. Please ensure that everything is well formatted and in a report style. You must also provide an explanation of the results in this section. 

Please ensure that everything is well labelled. So if you have multiple histograms and plots, calling them Figure 1, 2, 3, etc. and referencing them as Figure 1, Figure 2, etc. in your report will be expected. The reader should not get lost in a sea of information. Make sure to have the results be clean, well formatted and digestible.

# Discussion

Here you will summarize the previous sections and discuss conclusions drawn from the results. Make sure to elaborate and connect your analysis to the goal of the study.

## Weaknesses

Here we discuss weaknesses of the study, data, analysis, etc. You can also discuss areas for improvement.

## Next Steps

Here you discuss subsequent work to be done after this report. This can include next steps in terms of statistical analysis (perhaps there is a more efficient algorithm available, or perhaps there is a caveat in the data that would allow for some new technique). Future steps should also be specified in terms of the study setting (eg. including a follow-up survey on something, or a subsequent study that would complement the conclusions of your report).


# References



