---
title: "Estimating the result of 2020 US presidential election"
author: "Jessica Glustien, Nick Callow, Olivia Bi, Min Zhang"
date: "2 November 2020"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---


# Title of your Report

## Name(s) of Author(s) 
## Date


# Model



## Model Specifics

#Trump or not trump

```{r, include=FALSE}
library(tidyverse)
setwd("/Users/ZHANGMIN/Desktop")

survey_data <- read_csv("survey_data.csv")
census_data <- read_csv("census_data.csv")

# Creating the Model: Trump or not Trump
model_1 <- glm(vote_trump ~ age+as.factor(sex)+as.factor(race)+as.factor(education), 
            data=survey_data, family= "binomial")
summary(model_1)    # age >= 18

```

## Post-Stratification 

In order to estimate the proportion of voters who will vote for Donald Trump I need to perform a post-stratification analysis. Here I create cells based off different ages, race, eduction level. Using the model described in the previous sub-section I will estimate the proportion of voters in each age bin. I will then weight each proportion estimate (within each bin) by the respective population size of that bin and sum those values and divide that by the entire population size. 

```{r}
# Here I will perform the post-stratification calculation
census_data$logodds_estimate <-
  model_1 %>%
  predict(newdata = census_data)

census_data$estimate <-
  exp(census_data$logodds_estimate)/(1+exp(census_data$logodds_estimate))

census_data %>%
  mutate(trump_predict_prop = estimate*n) %>% 
  summarise(trump_predict = sum(trump_predict_prop)/sum(n))


```


#biden or not biden

```{r, include=FALSE}
library(tidyverse)
setwd("/Users/ZHANGMIN/Desktop")
survey_data <- read_csv("survey_data.csv")
census_data <- read_csv("census_data.csv")

# Creating the Model: Biden or not Biden
model_2 <- glm(vote_biden ~ age+as.factor(sex)+as.factor(race)+as.factor(education), 
            data=survey_data, family= "binomial")
summary(model_2)    # age >= 18
```

```{r}
census_data$logodds_est <-
  model_2 %>%
  predict(newdata = census_data)

census_data$estimate <-
  exp(census_data$logodds_est)/(1+exp(census_data$logodds_est))

model_2_result <-
  census_data %>% 
  mutate(biden_predict_prop = estimate*n) %>% 
  summarise(biden_predict = sum(biden_predict_prop)/sum(n))
model_2_result
```

#Trump or Biden: 

```{r,include=FALSE}
library(tidyverse)
setwd("/Users/ZHANGMIN/Desktop")
survey_data <- read_csv("survey_data_2.csv") # data only contain voters who will vote for Biden or Trump. 
census_data <- read_csv("census_data.csv")

# Creating the Model: Trump or Biden
model_3 <- glm(biden_or_trump ~ age+as.factor(sex)+as.factor(race)+as.factor(education), 
            data=survey_data, family= "binomial")
summary(model_3)    # age >= 18
```

```{r}
census_data$logodds_est <-
  model_3 %>%
  predict(newdata = census_data)

census_data$estimate <-
  exp(census_data$logodds_est)/(1+exp(census_data$logodds_est))

model_3_result <-
  census_data %>% 
  mutate(biden_win = estimate*n) %>% 
  summarise(biden_or_trump = sum(biden_win)/sum(n))
model_3_result
```

```{r, include=FALSE}
library(sjPlot)
library(sjmisc)
library(sjlabelled)
tab_model(model_1)
tab_model(model_2)
tab_model(model_3)
```

# Results

Based on three models above, we estimate that the proportion of voters in favor of voting for Joe Biden to be 0.507. This is based off our post-stratification analysis of the proportion of voters in favor of Joe Biden modelled by a logistic regression model, which accounted for age, sex, race and education.


# Discussion

Here you will summarize the previous sections and discuss conclusions drawn from the results. Make sure to elaborate and connect your analysis to the goal of the study.

## Weaknesses

Here we discuss weaknesses of the study, data, analysis, etc. You can also discuss areas for improvement.

## Next Steps

Here you discuss subsequent work to be done after this report. This can include next steps in terms of statistical analysis (perhaps there is a more efficient algorithm available, or perhaps there is a caveat in the data that would allow for some new technique). Future steps should also be specified in terms of the study setting (eg. including a follow-up survey on something, or a subsequent study that would complement the conclusions of your report).


# References



